Progress Report on the ``Adult Database'' Analysis

### 2009-11-12 report/paper.tex
Association rules and work plan.

\documentclass[a4paper]{llncs}
\usepackage{graphicx}
\usepackage{url}
\usepackage{longtable}
\usepackage[latin1]{inputenc}
\usepackage{subfigure}
\usepackage{a4wide}
\urldef{\mails}\path|{ei05011,ei05028}@fe.up.pt|
\bibliographystyle{splncs}
\begin{document}
\title{Progress Report on the ``Adult Database'' Analysis}
\author{Flávio Cruz \and João Azevedo}
\institute{Faculdade de Engenharia da Universidade do Porto\\
    Rua Dr. Roberto Frias, s/n 4200-465 Porto PORTUGAL\\
    \mails}
\maketitle

@@

Given a specific set of persons from a certain society and from a certain time we could then
learn what contributes to a person's social standing and, as we all know, money is an important
metric in that department. We could, for example, learn if the race affects the persons success and if
that society discriminates certain races.

Understanding what affects the money gained through an year can gives a lot of information
concerning society's behaviors.

@@

@@

@@

>>>>

\documentclass[a4paper]{llncs}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage[latin1]{inputenc}
\usepackage{subfigure}
\usepackage{a4wide}
\usepackage{listings}
\usepackage[hyphens]{url}
\urldef{\mails}\path|{ei05011,ei05028}@fe.up.pt|
\bibliographystyle{splncs}
\begin{document}
\title{Progress Report on the ``Adult Database'' Analysis}
\author{Flávio Cruz \and João Azevedo}
\institute{Faculdade de Engenharia da Universidade do Porto\\
    Rua Dr. Roberto Frias, s/n 4200-465 Porto PORTUGAL\\
    \mails}
\maketitle

@@

Given a specific set of persons from a certain society and from a certain time we could then
learn what contributes to a person's social standing and, as we all know, income is an important
metric in that department. We could, for example, learn if the race affects the persons success and if
that society discriminates against certain races.

Understanding what affects the income can give a lot of information concerning society's behaviors.

@@

Using association rules one can find frequent patterns and associations from a
set of transactions. These rules can easily be used to make data analysis and
to understand casual structures.

To build a set of association rules we used the apriori algorithm that is featured
in the R arules package. Before running the algorithm we made some data preprocessing
for continuous attributes that are needed to build better and more general rules.

The following attributes were transformed:

\begin{itemize}
  \item {\textbf{Age}: turned into categories. 15-25 (young), 25-45 (middle\_aged), 45-65 (senior), $>$ 65 (old).}
  \item {\textbf{Hours-per-week}: turned into categories. 0-25 (part\_time), 25-40 (full\_time), 40-60 (overtime), $>$60 (workaholic).}
  \item {\textbf{Capital-gain}: turned into categories. $<=$ 0 (none), $<$ median value (low), $>$ median value (high).}
  \item {\textbf{Capital-loss}: turned into categories. $<=$ 0 (none), $<$ median value (low), $>$ median value (high).}
\end{itemize}

Next, we plotted the relative frequency of items in the transaction database
with support greater than 10\%. The results are shown in figure \ref{fig:arules_item_frequency}.
We can learn various things from that plot: the majority of people are from the united states,
the white race is predominant, males outnumber females and that more people work in full time.

Finally we run the apriori algorithm using the test data set to generate rules
with support greater than 2\% and at least 60\% of confidence. The algorithm built
50487 rules.

We then built two sets of rules from the original set: one with the consequent component
with plus\_50=0 (the person gains less than 50K a year) and the other with plus\_50=1
(the person gains more than 50K a year).

Each rule set was sorted by the confidence level, from more confident to less confident.

The first 5 rules from the plus\_50=0 set are shown in listing \ref{arules_small_income} and
are related to small incomes. We can easily make some observations
about those rules: young people with a part-time are probably low income and having a child
without being married is also not a very good signal.

The second set, contrary to the ones above, shows us the top 5 rules for having large incomes.
These rules are shown in listing \ref{arules_large_income}. A conclusion can readily be made:
being a white husband with a high capital gain probably means we have a large income.

@@

Given the problem at hand, we want to generate a classification mechanism
that can tell if a specific person has a small or large income.

For this we will use various kinds of classification algorithms, from classical
decision trees to rule induction.

We will try to compare the error rate among the various algorithms. We'll be using
the train set to "train" the algorithm and then the test set to assess which algorithm
is best suited to make income predictions.

We'll also use some data-processing techniques to eliminate unknown values. It will
also be useful to categorize continuous values for more generalized rules as we have
done when generating association rules.

The algorithms we will be using are: ID3, C4.5, CN2 and AQ. Depending on our progress
we will experiment with neural networks and genetic algorithms, which are other
mechanisms for building classifiers.

@@

\begin{figure}
\centering
\includegraphics[width=20cm,angle=90]{item_frequency.png}
\caption{Item frequencies of items with at least 10\% support.}
\label{fig:arules_item_frequency}
\end{figure}

\clearpage
\section{Listings}

\begin{lstlisting}[language=r,frame=single,breaklines=true,basicstyle=\footnotesize\ttfamily,caption={Association rules for small income.},label=arules_small_income]
  lhs                               rhs            support confidence     lift
1 {age=young,                                                                 
   hours_per_week=part_time}     => {plus_50=0} 0.05730782          1 1.317193
2 {age=young,                                                                 
   occupation=unknown,                                                        
   capital_gain=none}            => {plus_50=0} 0.02002396          1 1.317193
3 {age=young,                                                                 
   relationship=own_child,                                                    
   hours_per_week=part_time}     => {plus_50=0} 0.04404042          1 1.317193
4 {education=some_college,                                                    
   relationship=own_child,                                                    
   hours_per_week=part_time}     => {plus_50=0} 0.02069961          1 1.317193
5 {marital_status=never_married,                                              
   relationship=own_child,                                                    
   hours_per_week=part_time}     => {plus_50=0} 0.04689659          1 1.317193
\end{lstlisting}

\begin{lstlisting}[language=r,frame=single,breaklines=true,basicstyle=\footnotesize\ttfamily,caption={Association rules for large income.},label=arules_large_income]
  lhs                                    rhs            support confidence     lift
1 {relationship=husband,                                                           
   race=white,                                                                     
   capital_gain=high}                 => {plus_50=1} 0.02063819  0.9926145 4.121990
2 {marital_status=married_civ_spouse,                                              
   relationship=husband,                                                           
   race=white,                                                                     
   capital_gain=high}                 => {plus_50=1} 0.02063819  0.9926145 4.121990
3 {relationship=husband,                                                           
   race=white,                                                                     
   sex=male,                                                                       
   capital_gain=high}                 => {plus_50=1} 0.02063819  0.9926145 4.121990
4 {relationship=husband,                                                           
   race=white,                                                                     
   capital_gain=high,                                                              
   capital_loss=none}                 => {plus_50=1} 0.02063819  0.9926145 4.121990
5 {relationship=husband,                                                           
   capital_gain=high}                 => {plus_50=1} 0.02251159  0.9918809 4.118943
\end{lstlisting}
